---
title: "Randomized Controlled Trials"
author: "Nicholas Skelley"
date: "4/6/2020"
output: 
  html_document: 
    highlight: breezedark
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

####################################################################
## Packages
# To install and open the R packages that you need for this code. 
need <- c('tidyverse','kableExtra','glue','stargazer', 'foreign', 'broom')
have <- need %in% rownames(installed.packages()) 
if(any(!have)) install.packages(need[!have]) 
invisible(lapply(need, library, character.only=T)) 

# Change path to whereever you place the models
# To set up the working directory. 
rm(list = ls())
setwd(getwd())
####################################################################
```

<style type="text/css">
body{ /* Normal  */
    font-size: 16px;
}
h1.title {
  font-size: 38px;
}
h1 { /* Header 1 */
  font-size: 25px;
}
h2 { /* Header 2 */
  font-size: 22px;
}
</style>

Read the paper “The Effects of Micro-entrepreneurship Programs on Labor Market Performance: Experimental Evidence from Chile” by Claudia Martínez A., Esteban Puentes, Jaime Ruiz-Tagle. The authors are interested in whether a training program in Chile increased self-employment. In particular, pay attention to the motivation fo the question and the context/experimental setup and the details of the program they are evaluating.

```{r}
dataset <- read.dta("data/RCT_Martinezetal.dta")

dataset <- dataset %>%
  filter(anho==2011)
```

**Show that covariates are balanced across treatment status. These variables include sex (`sexo2010`), age (`edad2010`),  education==primary school (`basica2010`), working at baseline (`trabaja_base`), wage working at baseline (`dependiente_base`). Generate a table (or some form of output) that contains the results of a statistical test (t-test of means across groups) for these variables.**

```{r}
dataset$male <- dataset$sexo2010=="hombre"

# table of means
dataset %>%
  group_by(T3) %>%
  select(T3, male, edad2010, basica2010, trabaja_base, dependiente_base) %>%
  summarise_all(mean, na.rm=T) %>%
  select(male, edad2010, basica2010, trabaja_base, dependiente_base) %>%
  t() %>%
  kable(col.names = c("Control","Treated"), digits=3) %>%
  kable_styling(full_width = F)
```

```{r, results = "asis"}
# table of differences in means
stargazer(list(lm(male~T3, dataset), lm(edad2010~T3, dataset), lm(basica2010~T3, dataset), lm(trabaja_base~T3, dataset), lm(dependiente_base~T3, dataset)), type="html", omit=c("Constant"), keep.stat=c("n"), covariate.labels = c("Treated"), dep.var.labels=c("Male","Age","Primary School","Working","Wage"))
```

```{r, results = "hold"}
# series of t-tests
t.test(male ~ T3, dataset)
t.test(edad2010 ~ T3, dataset)
t.test(basica2010 ~ T3, dataset)
t.test(trabaja_base ~ T3, dataset)
t.test(dependiente_base ~ T3, dataset)
```



**Present a table of means of the variable for being self-employed in both periods (`indep_indep`) across the two (treament & control) groups.**

```{r}
dataset %>%
  group_by(T3) %>%
  summarise(mean_indep_indep=mean(indep_indep, na.rm=T)) %>%
  select(mean_indep_indep) %>%
  t() %>%
  kable(digits=3, col.names=c("Control","Treated")) %>%
  kable_styling(full_width=F)
```

**Write the regression equation that you will estimate using linear regression to get the effect of the program on employment outcomes in this RCT.**

$$indep\_indep_i = \beta_0 + \beta_1 \cdot Treated_i + u_i$$
```{r}
model1 <- lm(indep_indep~T3, dataset)
```


**What is the effect of treatment on being self-employed in both periods (indep_indep)? Determine some of the covariates you could possibly control for (that are in the dataset) and add them to the regression specification.**

```{r}
summary(model1)
# receiving treatment increases the probability of being self-employed in both periods by 13.3 percentage points
```

$$indep\_indep_i = \beta_0 + {\beta_1 \cdot Treated_i} + {\beta_2 \cdot PrimSchool_i} + {\beta_3 \cdot Skill_i} + {\beta_4 \cdot Age_i} + {\beta_5 \cdot Male_i} + {\beta_6 \cdot Working_i} + {\beta_7 \cdot Wage_i} + u_i$$

```{r}
model2 <- lm(indep_indep ~ T3 + basica2010 + skills_index + edad2010 + male + trabaja_base + dependiente_base, dataset)
summary(model2)
# the estimate for the effect of treatment on the probability of being self-employed in both periods decreases by about one percentage point. The standard error of beta1_hat is slightly lower in the expanded regression.

dataset %>%
  filter(T3==0 & indep_indep==1) %>%
  nrow()
# 96 people in the control group are self-employed in both periods

# there are 172 NA values of indep_indep in the control group; this accounts for the low number of people in the control group who are self-employed in both periods
dataset %>%
  filter(T3==0 & is.na(indep_indep)) %>%
  nrow()
```

<br>