---
title: "Regression Discontinuity"
author: "Nicholas Skelley"
date: "4/13/2020"
output: 
  html_document: 
    highlight: breezedark
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

####################################################################
## Packages
# To install and open the R packages that you need for this code. 
need <- c('tidyverse','reshape2','kableExtra','glue','stargazer','foreign','plyr')
have <- need %in% rownames(installed.packages()) 
if(any(!have)) install.packages(need[!have]) 
invisible(lapply(need, library, character.only=T)) 

# Change path to whereever you place the models
# To set up the working directory. 
rm(list = ls())
setwd(getwd())
####################################################################
```

<style type="text/css">
body{ /* Normal  */
    font-size: 16px;
}
h1.title {
  font-size: 38px;
}
h1 { /* Header 1 */
  font-size: 25px;
}
h2 { /* Header 2 */
  font-size: 22px;
}
</style>

The goal of this problem is to work through the construction of an RD paper. For our example, we will
use Lee (2008). There are two datasets: `Lee2008Candidate.dta` and `Lee2008Party.dta`.

```{r results='hide'}
party <- read.dta("data/Lee2008Party.dta") %>% filter(use==1)
candidate <- read.dta("data/Lee2008Candidate.dta") %>% filter(use==1)

# candidate$second --> finished second in the race indicator (if they won the election, didn't run, or finished below second, value is NA); (NOT about the *next* race)

candidate$second[is.na(candidate$second)] <- 0

# naive comparison (probably incorrect)
mean(candidate$myoutcomenext[candidate$myoutcome==1],na.rm=T)-mean(candidate$myoutcomenext[candidate$myoutcome==0],na.rm=T)

nextoutcome <- melt(merge(candidate %>%
                      group_by(yearel, myoutcome) %>%
                      dplyr::summarise(avg_nextres=mean(myoutcomenext, na.rm=T)) %>%
                      dcast(yearel~myoutcome, value.var="avg_nextres"),
                    party %>%
                      group_by(yearel) %>%
                      dplyr::summarise(avg_repeat=mean(repeatparty, na.rm=T))), 
                    variable.name="lastres", value.name="nextoutcome", id.vars=c("yearel")) %>%
                mutate(lastres=as.character(lastres)) %>%
                mutate(lastres=ifelse(lastres=="0", "runnerup", ifelse(lastres=="1", "winner", "partywon")))

```

### Replicate Figure 1.

```{r}
ggplot(nextoutcome, aes(x=yearel, y=nextoutcome, group=lastres, 
                        linetype=lastres, shape=lastres)) + 
  geom_line() +
  geom_point(size=2.5) +
  scale_linetype_manual(values=c("solid","dashed","dashed"), 
                        labels=c("Incumbent Party", "Runner-up Candidate", 
                                 "Winning Candidate")) +
  scale_shape_manual(values=c(15,4,0), labels=c("Incumbent Party", 
                                                "Runner-up Candidate",
                                                "Winning Candidate")) +
  labs(x="Year",y="Proportion Winning Election") +
  theme(legend.title=element_blank())
```

### Replicate Figure 2a three times, with binwidths of 0.02, 0.01, and 0.005.

```{r, warning=F}
# using candidate
candidate %>%
  group_by(roundedDiff=round_any(difshare, 0.02)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2.5) + 
  xlim(-0.25,0.25) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Figure 2a, binwidth=0.02")

candidate %>%
  group_by(roundedDiff=round_any(difshare, 0.01)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2.5) + 
  xlim(-0.25,0.25) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Figure 2a, binwidth=0.01")

candidate %>%
  group_by(roundedDiff=round_any(difshare, 0.005)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2.5) + 
  xlim(-0.25,0.25) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Figure 2a, binwidth=0.005")
```

### Write the regression equation that you will estimate.

$$ prob\_wins_i = \beta_0 + \beta_1 \cdot won\_last_i + u_i $$

We now want to estimate the ATE of incumbency on vote share in the next election. Construct a column
with four estimates and a bandwidth of 0.25. Column 1 has local linear regression, and Columns 2-4 have
2nd-4th order polynomial controls in the forcing variable. Make sure to estimate separate polynomials on
either side of the cutoff.

#### Model 2
$$ prob\_wins_i = \beta_0 + \beta_1 \cdot won\_last_i + \beta_2 \cdot victory\_margin_i + u_i $$

#### Model 3
$$ prob\_wins_i = \beta_0 + \beta_1 \cdot won\_last_i + \beta_2 \cdot victory\_margin_i + \beta_3 \cdot victory\_margin_i^2 + u_i $$

#### Model 4
$$ prob\_wins_i = \beta_0 + \beta_1 \cdot won\_last_i + \beta_2 \cdot victory\_margin_i + \beta_3 \cdot victory\_margin_i^2 + \beta_4 \cdot won\_last_i \cdot victory\_margin_i^2 + u_i $$

```{r, results="asis", warning=F}
candidate.rd <- candidate %>%
  filter(abs(difshare) <= 0.25)

model1 <- lm(myoutcomenext~myoutcome, candidate.rd)
model2 <- lm(myoutcomenext~myoutcome+difshare, candidate.rd)
model3 <- lm(myoutcomenext~myoutcome+difshare+I(difshare^2), candidate.rd)
model4 <- lm(myoutcomenext~myoutcome+difshare+I(difshare^2)+I(difshare^2)*myoutcome, candidate.rd)

#graph of model1
candidate.rd %>%
  group_by(roundedDiff=round_any(difshare,0.005)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext,na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2) + 
  xlim(-0.25,0.25) + 
  ylim(0,1) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x<=0, coef(model1)[1], NA)}) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x>=0, coef(model1)[1] + coef(model1)[2], NA)}) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Model 1")
  
#graph of model2
candidate.rd %>%
  group_by(roundedDiff=round_any(difshare, 0.005)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2) + 
  xlim(-0.25,0.25) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x<=0, coef(model2)[1] + (coef(model2)[3] * x), NA)}) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x>=0, coef(model2)[1] + (coef(model2)[3] * x) + coef(model2)[2], NA)}) + 
  ylim(0,1) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Model 2")

#graph of model3
candidate.rd %>%
  group_by(roundedDiff=round_any(difshare, 0.005)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2) + 
  xlim(-0.25,0.25) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x<=0, coef(model3)[1] + (coef(model3)[3] * x) + (coef(model3)[4] * (x^2)), NA)}) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x>=0, coef(model3)[1] + (coef(model3)[3] * x) + (coef(model3)[4] * (x^2)) + coef(model3)[2], NA)}) + 
  ylim(0,1) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Model 3")

# graph of model4
candidate.rd %>%
  group_by(roundedDiff=round_any(difshare, 0.005)) %>%
  dplyr::summarise(prop_reelect=mean(myoutcomenext, na.rm=T)) %>%
  ggplot(aes(x=roundedDiff, y=prop_reelect)) + 
  geom_point(size=2) + 
  xlim(-0.25,0.25) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x<=0, coef(model4)[1] + (coef(model4)[3] * x) + (coef(model4)[4] * x^2), NA)}) +
  stat_function(color="red", size=1, fun=function(x){ifelse(x>=0, coef(model4)[1] + coef(model4)[2] + (coef(model4)[3] * x) + ((coef(model4)[4] + coef(model4)[5]) * x^2), NA)}) + ylim(0,1) +
  labs(x="Margin of Victory, Election t", 
       y="Probability of Victory, Election t+1", 
       title="Model 4")

stargazer(list(model1,model2,model3,model4), type="html", dep.var.labels = "Probability of Victory, Election t+1", covariate.labels = c("Incumbent","Margin of Victory","Margin of Victory Sq.","Incumbent * Margin of Victory Sq."))
```

<br>
